---
title: "Prognostic Score Construction and MAP Prior Workflow"
author: "Emmanuel"
format: html
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Overview

This document describes The Data generation procedure and the Bayesian
Borrowing steps:

# Data Generation

The goal is to create a realistic clinical trial setting with:

-   Multiple **historical studies** (placebo only),
-   One **current randomized trial** (placebo + treatment),
-   High-dimensional, collinear covariates,
-   Both **prognostic** and **predictive** covariate effects,
-   Optional **distributional drift** between historical and current
    data.

The full data-generating process is known and controlled, enabling
principled evaluation of borrowing and transfer-learning methods.

------------------------------------------------------------------------

## Global Simulation Parameters

```{r}
library(MASS)

n_hist <- 200      # Patients per historical study
n_trial <- 500     # Patients in current trial
H <- 5             # Number of historical studies
p_cont <- 50       # Number of continuous features
p_bin <- 5         # Number of binary biomarkers
rho <- 0.8         # Correlation for collinearity
global_hist_drift <- 0  # Global historical vs current shift

```

As an example (just to show for a replicate)

The simulation consists of:

-   $H = 5$ historical studies, each with $n_{hist}=200$ subjects,
-   One current study with $n_{trial} =500$ subjects,
-   $p=55$ covariates in total:
    -   $50$ continuous laboratory variables,
    -   $5$ binary biomarkers.

## Covariates Generation

### Continuous Features

Continuous covariates are generated from a multivarite normal
distribution with strong colinearity. The covariance matrix follows the
[Toeplitz](https://en.wikipedia.org/wiki/Toeplitz_matrix "Toeplitz matrix")
structure:

$$
\Sigma_{ij}=\rho^{|i-j|}, \ \ \ \ \rho = 0.8
$$
This is made to include realistic correlation patterns commonly observed in clinical laboratory data.

### Binary Biomarkers

Binary biomarkers represent genetic or molecular indicators and are generated independently from a Bernolli distribution with success probability 0.3.

The procedure for the covariates generation is implemented in the following function:

```{r}

generate_x <- function(n) {
  sigma <- toeplitz(rho^(0:(p_cont - 1)))
  X_cont <- mvrnorm(n, mu = rep(0, p_cont), Sigma = sigma)
  X_bin <- matrix(rbinom(n * p_bin, 1, 0.3), nrow = n)

  colnames(X_cont) <- paste0("Lab_", 1:p_cont)
  colnames(X_bin) <- paste0("Bio_", 1:p_bin)

  cbind(X_cont, X_bin)
}



```


## Outcome Model

The continuous outcome variable $Y$ is generated as the sum of:

1. Prognostic signal (covariate main effects),

2. Predictive treatment effect (treatment–biomarker interaction),

3. Study-level drift (historical studies only),

4. Global historical drift (optional, a measure of descrepancy between the joint historical data and the current data),

5. Random noise.

### Prognostic component

The prognostic signal is nonlinear and involves interactions:

$$
f(x) = 2Lab_1 +3Lab_2^2 +5sign(Lab_3) + 4Lab_4 \times Lab_5
$$

### Predictive (Treatment) component

Treatment effects depend on biomarker status:

$$
g(X,T)=5T+12T\times Bio_1
$$
This creates a strong predictive biomarker, where subjects with $Bio_1=1$ experuence substantially larger treatment effects.

### Noise and Drift

- Observation-level noise: $\varepsilon \sim N(0,2^2)$
- Inter study drift: obtained from a random Uniform sample $[-2,2]$ (historical study only)
- Global historical drift: constant shift applied to all historical data, to shift it from the current trial.

The function for the outcome generation is as follows:

```{r}
generate_y <- function(X, Trt,
                       study_drift = 0,
                       global_hist_drift = 0) {

  prog_signal <- (2 * X[, "Lab_1"]) +
    (3 * X[, "Lab_2"]^2) +
    (5 * sin(X[, "Lab_3"])) +
    (4 * X[, "Lab_4"] * X[, "Lab_5"])

  trt_effect <- (5 * Trt) + (12 * Trt * X[, "Bio_1"])

  noise <- rnorm(nrow(X), mean = 0, sd = 2)

  prog_signal +
    trt_effect +
    study_drift +
    global_hist_drift +
    noise
}


```


## Historical Data Generation

Each historical study:

- Contains placebo-only subjects,

- Shares the same covariate distribution,

- Has its own random study-level drift,

- Optionally shares a pooled historical drift.

```{r}
historical_data <- lapply(1:H, function(s) {
  X <- generate_x(n_hist)
  Trt <- rep(0, n_hist)

  Y <- generate_y(
    X, Trt,
    study_drift = runif(1, -2, 2),
    global_hist_drift = global_hist_drift
  )

  data.frame(
    Study = paste0("H", s),
    Treatment = Trt,
    Target = Y,
    X
  )
})


```




## Current Trial Generation

The current trial differs from historical data in that:

- Subjects are randomized to treatment with probability 0.7,

- No study-level or global drift is applied,

- Both prognostic and predictive effects are active

```{r}
X_trial <- generate_x(n_trial)
Trt_trial <- rbinom(n_trial, 1, 0.7)
Y_trial <- generate_y(X_trial, Trt_trial)

current_trial <- data.frame(
  Study = "Current",
  Treatment = Trt_trial,
  Target = Y_trial,
  X_trial
)

all_data <- do.call(rbind, c(historical_data, list(current_trial)))


```



## Just to check the predictive Biomarker Effect

```{r}
library(ggplot2)

ggplot(
  current_trial,
  aes(x = factor(Bio_1), y = Target, fill = factor(Treatment))
) +
  geom_boxplot() +
  labs(
    title = "Treatment Effect Modulated by Bio_1",
    x = "Biomarker 1 Status",
    fill = "Treatment"
  ) +
  theme_minimal()


```

# Bayesian Borrowing technique

1.  Construction of a **prognostic score** using historical control
    data.
2.  Construction of a **MAP (Meta-Analytic-Predictive) prior** using the
    prognostic score.
3.  Application of the MAP prior in the current trial.

Assume:

-   $K$ historical trials
-   Patient-level covariates $X_i$
-   Outcome $Y_i$
-   Treatment indicator $w_i$

------------------------------------------------------------------------

# Notation

For patient $i$ in study $k$:

$$
(X_{ik}, Y_{ik}, w_{ik})
$$

Where:

-   $X_{ik} \in \mathbb{R}^p$: baseline covariates
-   $Y_{ik}$: outcome
-   $w_{ik} \in \{0,1\}$: treatment indicator

Let:

$$
\mathcal{H} = \bigcup_{k=1}^K D_k^{(control)}
$$

be the pooled historical control dataset.

------------------------------------------------------------------------

# Step 1: Build the prognostic model

The prognostic score is defined as:

$$
m_i = E(Y_i \mid X_i, \text{Control})
$$

We estimate this function using a random forest trained on historical
controls.

Formally:

$$
Y_i^{H} = f(X_i^{H}) + \varepsilon_i
$$

where:

-   $f(\cdot)$ is an unknown nonlinear function
-   estimated using any machine learning model.

The prognostic score is:

$$
m_i = \hat{f}(X_i)
$$

## R: Fit Random Forest Prognostic Model

```{r}
library(randomForest)


```

### Important note on drift:

There are two choices on how to fuit the prognostic models to/not to
capture the inter-trial variability:

Option A: Ignore the study indicator (this is simple pooling of the
historical data)

$$Y \sim X$$ This assumes that the prognostic relationship is similar
across trials.

Option B: Include the study as a feature of the prognostic model:

$$
Y \sim X + Study
$$ This captures the inter-trial differences and baseline shifts.

------------------------------------------------------------------------

# Step 2: Compute the prognostic scores

Once the Prognostic model is built (trained):

Compute the prognostic scores as follows:

-   For historical Controls:\
    $$
    m_i^H = \hat{f}\left(X_i^H\right)
    $$
-   For current trial: $$
    m_i^C = \hat{f}\left(X_i^C\right)
    $$

So that every patient (regardless of where s(he) belongs) has $m_i$,
\[prognostic score\]

------------------------------------------------------------------------

# Step 3 Build the MAP prior (Informative component)

Now ignore the original covariates and use only the outcome $Y$ and the
prognostic score $m$ for the historical controls.

**For each dataset** $k$

Fit: $$
Y_{ik} = \beta_{0k} + \beta_{2k}m_{ik} + \varepsilon_{ik}
$$ This gives the posterior:
$\pi_k\left(\beta_0, \beta_2,\sigma^2\right)$

These are combined across studies using MAP to get the $MAP$ prior. And
can be converted to a mixture of conjugate distributions
using`automixfit` function from the `RBesT` package.

------------------------------------------------------------------------

# Step 4: "Robustification"

The prior can then be robustified using the `RBesT` package to get the
informative and the non informative components of the prior. Where an
initial weight of the robust component is given as an argument of the
function.

------------------------------------------------------------------------

# Step 5: Analysis Model in the Current trial

Now, the analysis model is fitted:

$$
Y_i = \beta_0 + \beta_1w_i + \beta_2m_i + varepsilon_i
$$

With the mixture prior:

$$
\pi(\theta) = \omega \pi_1(\theta) + (1-\omega)\pi_0(\theta)
$$

where: $\pi_1$ is the informative prior (from the historical data)
$\pi_0$ is the non informative (Vague) component.
